---
title: "Multiobjective Analysis Code"
format: html
editor: visual
---

# Packages and Libraries

```{r}
library(writexl)
library(gsheet)
library(r4pde)
library(tidyverse)
library(ggdendro)
library(ggrepel)
library(cluster)
library(mclust)   # adjustedRandIndex
library(patchwork)
```

# Data Loading

```{r}
dat_raw <- gsheet2tbl("https://docs.google.com/spreadsheets/d/1KGolCesLlgn2TGS-YcuKtdqNVJ4OnkdG/edit?gid=1573255184#gid=1573255184")
dat_raw <- dat_raw %>%
  mutate(Genotype = recode(Genotype,
                           "BRS Gralha Azul" = "BRS GAzul",
                           "ORS Madreperola" = "ORS MPerola"))
```

# Data Cleaning and Preparation

```{r}
dat_all <- dat_raw %>%
  mutate(
    geno = str_squish(as.character(Genotype)),
    env  = str_squish(as.character(Env)),
    time = as.numeric(Zadoks),
    tan_spot  = as.numeric(Tan_spot),
    rust = as.numeric(rust),
    powdery_mildew = as.numeric(powdery_mildew)) %>%
  filter(
    geno %in% Genotype,
    !is.na(env), env != "",
    !is.na(time),
    !is.na(tan_spot),
    !is.na(rust),
    !is.na(powdery_mildew),) %>%
  arrange(env, geno, time)
```

# Filtering Genotypes Evaluated Across Multiple Environments

```{r}
filtered_data <- dat_all %>%
  #filter(!is.na(time)) %>%
  #group_by(geno, env) %>%
  #filter(n_distinct(time) >= 5) %>%
  #ungroup() %>%
  group_by(geno) %>%
  filter(
    n_distinct(Local) >= 4,
    n_distinct(Year)  >= 3) %>%
  ungroup()


filtered_data %>%
  distinct(geno) %>%
  arrange(geno) |> 
  print(n = Inf)
n_distinct(filtered_data$geno)


genotype_env_count <- filtered_data %>%
  group_by(geno) %>%
  summarise(
    n_env   = n_distinct(env),
    n_year  = n_distinct(Year),
    n_local = n_distinct(Local),
    .groups = "drop") %>%
  arrange(desc(n_env))

genotype_env_count
#write_xlsx(genotype_env_count, "genotype_environment_count.xlsx")
```

# Filtering Environments with Final Severity \> 10%

```{r}
# PM
final_sev_pm <- filtered_data %>%
  group_by(env, geno) %>%
  filter(time == max(time, na.rm = TRUE)) %>%
  summarise(final_sev = max(powdery_mildew, na.rm = TRUE), .groups = "drop")

env_valid_pm <- final_sev_pm %>%
  group_by(env) %>%
  summarise(max_final = max(final_sev, na.rm = TRUE), .groups = "drop") %>%
  filter(max_final > 10)

dat_env_pm <- filtered_data %>%
  filter(env %in% env_valid_pm$env) %>%
  arrange(env, geno, time)


# TS
final_sev_ts <- filtered_data %>%
  group_by(env, geno) %>%
  filter(time == max(time, na.rm = TRUE)) %>%
  summarise(final_sev = max(Tan_spot, na.rm = TRUE), .groups = "drop")

env_valid_ts <- final_sev_ts %>%
  group_by(env) %>%
  summarise(max_final = max(final_sev, na.rm = TRUE), .groups = "drop") %>%
  filter(max_final > 10)

dat_env_ts <- filtered_data %>%
  filter(env %in% env_valid_ts$env) %>%
  arrange(env, geno, time)


# Rust
final_sev_rust <- filtered_data %>%
  group_by(env, geno) %>%
  filter(time == max(time, na.rm = TRUE)) %>%
  summarise(final_sev = max(rust, na.rm = TRUE), .groups = "drop")

env_valid_rust <- final_sev_rust %>%
  group_by(env) %>%
  summarise(max_final = max(final_sev, na.rm = TRUE), .groups = "drop") %>%
  filter(max_final > 10)

dat_env_rust <- filtered_data %>%
  filter(env %in% env_valid_rust$env) %>%
  arrange(env, geno, time)

```

# Breeder Resistance Classes (Genotype-level)

```{r}
resistencia_tbl <- tribble(
  ~geno,            ~rust_cl, ~ts_cl, ~pm_cl,
  "Bar 10",         "MR",     "MR",   "MS",
  "BRS GAzul",      "R",      "AS",   "R",
  "BRS Sanhaco",    "R",      "AS",   "R",
  "BRS TR133",      "R",      "AS",   "R",
  "BRS TR733",      "R",      "AS",   "R",
  "ORS 1403",       "MS",     "MR",   "R",
  "ORS Feroz",      "R",      "MS",   "MS",
  "ORS MPerola",    "R",      "MR",   "AS",
  "ORS Selvagem",   "R",      "MS",   "R",
  "ORS Senna",      "MR",     "AS",   "R",
  "ORS Soberano",   "MS",     "MS",   "MS",
  "ORS Turbo",      "MR",     "MR",   "MS",
  "TBIO Aton",      "MR",     "AS",   "R",
  "TBIO Audaz",     "MR",     "MR",   "AS",
  "TBIO Blanc",     "AS",     "MR",   "MS",
  "TBIO Calibre",   "R",      "MS",   "MS",
  "TBIO Capaz",     "MR",     "MR",   "AS",
  "TBIO Duque",     "AS",     "MS",   "AS",
  "TBIO Motriz",    "MR",     "MR",   "MS",
  "TBIO Ponteiro",  "AS",     "MS",   "R",
  "TBIO Sossego",   "MS",     "MR",   "MS",
  "TBIO Talisma",   "MR",     "MR",   "MR",
  "TBIO Titan",     "MS",     "MS",   "MR",
  "TBIO Toruk",     "MR",     "MS",   "MR",
  "TBIO Trunfo",    "MS",     "MS",   "MS"
) %>%
  mutate(geno = str_squish(geno))
cultivars <- resistencia_tbl$geno

geno_class <- resistencia_tbl %>%
  mutate(
    geno = str_squish(geno),
    pm_cl = factor(pm_cl, levels = c("R", "MR", "MS", "AS"), ordered = TRUE))
```

# Map Resistance Classes to Numeric Values

```{r}
map_res <- c(
  "R" = 1,
  "MR" = 2,
  "MS" = 3,
  "AS" = 4)

cl_rust   <- setNames(unname(map_res[resistencia_tbl$rust_cl]), resistencia_tbl$geno)
cl_mildew <- setNames(unname(map_res[resistencia_tbl$pm_cl]), resistencia_tbl$geno)
cl_tan    <- setNames(unname(map_res[resistencia_tbl$ts_cl]), resistencia_tbl$geno)


cl_rust <- setNames(
  unname(map_res[resistencia_tbl$rust_cl]),
  resistencia_tbl$geno)

cl_mildew <- setNames(
  unname(map_res[resistencia_tbl$pm_cl]),
  resistencia_tbl$geno)

cl_tan <- setNames(
  unname(map_res[resistencia_tbl$ts_cl]),
  resistencia_tbl$geno)

```

# Retaining Selected Genotypes and Joining Breeder Classes

```{r}
#ts
dat_env_ts  <- dat_env_ts  %>%
  mutate(geno = str_squish(geno))

resistencia_tbl <- resistencia_tbl %>%
  mutate(geno = str_squish(geno))

dat_env_ts  <- dat_env_ts  %>%
  filter(geno %in% resistencia_tbl$geno)

dat_env_ts  <- dat_env_ts  %>%
  left_join(resistencia_tbl, by = "geno")

n_distinct(dat_env_ts$geno)

#Rust

dat_env_rust   <- dat_env_rust   %>%
  mutate(geno = str_squish(geno))

resistencia_tbl <- resistencia_tbl %>%
  mutate(geno = str_squish(geno))

dat_env_rust   <- dat_env_rust   %>%
  filter(geno %in% resistencia_tbl$geno)

dat_env_rust   <- dat_env_rust   %>%
  left_join(resistencia_tbl, by = "geno")

n_distinct(dat_env_rust$geno)

# PM
dat_env_pm<- dat_env_pm    %>%
  mutate(geno = str_squish(geno))

dat_env_pm<- dat_env_pm  %>%
  mutate(geno = str_squish(geno))

dat_env_pm<- dat_env_pm    %>%
  filter(geno %in% resistencia_tbl$geno)

dat_env_pm<- dat_env_pm    %>%
  left_join(resistencia_tbl, by = "geno")

n_distinct(dat_env_pm$geno)

```

# Aligning Disease Progress Curves to Zadoks Growth Scale

```{r}
#ts

dae_min <- 20
dae_max <- 90

starts <- dat_env_ts %>%
  filter(between(time, dae_min, dae_max)) %>%
  group_by(env, geno) %>%
  summarise(
    t_pos = {
      v <- time[Tan_spot > 0]
      if (length(v) == 0) NA_real_ else min(v)},
    t_zero_prev = {
      if (is.na(t_pos)) NA_real_ else {
        v0 <- time[time < t_pos]
        if (length(v0) == 0) NA_real_ else max(v0)
      }
    },
    t_start = dplyr::coalesce(t_zero_prev, t_pos),
    .groups = "drop")

dat_trim <- dat_env_ts %>%
  filter(between(time, dae_min, dae_max)) %>%
  left_join(starts %>% dplyr::select(env, geno, t_start), by = c("env", "geno")) %>%
  filter(is.na(t_start) | time >= t_start) %>%
  dplyr::select(-t_start) %>%
  arrange(env, geno, time)

# Convert to proportion for compare_curves()
dat_ready1 <- dat_trim %>%
  mutate(
         yts = Tan_spot/100)
n <- nrow(dat_ready1)

dat_ready1$yts <- (dat_ready1$yts * (n - 1) + 0.5) / n
################################Rust########################
dae_min <- 20
dae_max <- 90

starts <- dat_env_rust %>%
  filter(between(time, dae_min, dae_max)) %>%
  group_by(env, geno) %>%
  summarise(
    t_pos = {
      v <- time[rust > 0]
      if (length(v) == 0) NA_real_ else min(v)},
    t_zero_prev = {
      if (is.na(t_pos)) NA_real_ else {
        v0 <- time[time < t_pos]
        if (length(v0) == 0) NA_real_ else max(v0)
      }
    },
    t_start = dplyr::coalesce(t_zero_prev, t_pos),
    .groups = "drop")

dat_trim <- dat_env_rust %>%
  filter(between(time, dae_min, dae_max)) %>%
  left_join(starts %>% dplyr::select(env, geno, t_start), by = c("env", "geno")) %>%
  filter(is.na(t_start) | time >= t_start) %>%
  dplyr::select(-t_start) %>%
  arrange(env, geno, time)

# Convert to proportion for compare_curves()
dat_ready2 <- dat_trim %>%
  mutate(
         yrust = rust/100)

n <- nrow(dat_ready2)

dat_ready2$yrust <- (dat_ready2$yrust * (n - 1) + 0.5) / n
# PM
dae_min <- 20
dae_max <- 90

starts <- dat_env_pm %>%
  filter(between(time, dae_min, dae_max)) %>%
  group_by(env, geno) %>%
  summarise(
    t_pos = {
      v <- time[powdery_mildew > 0]
      if (length(v) == 0) NA_real_ else min(v)},
    t_zero_prev = {
      if (is.na(t_pos)) NA_real_ else {
        v0 <- time[time < t_pos]
        if (length(v0) == 0) NA_real_ else max(v0)
      }
    },
    t_start = dplyr::coalesce(t_zero_prev, t_pos),
    .groups = "drop")

dat_trim <- dat_env_pm %>%
  filter(between(time, dae_min, dae_max)) %>%
  left_join(starts %>% dplyr::select(env, geno, t_start), by = c("env", "geno")) %>%
  filter(is.na(t_start) | time >= t_start) %>%
  dplyr::select(-t_start) %>%
  arrange(env, geno, time)

# Convert to proportion for compare_curves()
dat_ready3 <- dat_trim %>%
  mutate(
         ypm = powdery_mildew/100)
n <- nrow(dat_ready3)

dat_ready3$ypm <- (dat_ready3$ypm * (n - 1) + 0.5) / n
```

# Checking the Distribution of Classification

```{r}
#RUST
dat_ready2 %>%
  distinct(geno, rust_cl) %>%
  count(rust_cl)
#TS
dat_ready1 %>%
  distinct(geno, ts_cl) %>%
  count(ts_cl)
#pm
dat_ready3 %>%
  distinct(geno, pm_cl) %>%
  count(pm_cl)
```

# Hierarchical Generalized Additive Model (HGAM) Fitting

```{r}

dat_ready1 <- dat_ready1 %>%
  mutate(curve_id = interaction(env, geno, drop = TRUE))
dat_ready2 <- dat_ready2 %>%
  mutate(curve_id = interaction(env, geno, drop = TRUE))
dat_ready3 <- dat_ready3 %>%
  mutate(curve_id = interaction(env, geno, drop = TRUE))

library(nlme)
library(mgcv)

dat_ready1 <- dat_ready1 %>%
  mutate(
    env = factor(env),
    geno = factor(geno),
    curve_id = factor(curve_id),
    time = as.numeric(time))

dat_ready2 <- dat_ready2 %>%
  mutate(
    env = factor(env),
    geno = factor(geno),
    curve_id = factor(curve_id),
    time = as.numeric(time))

dat_ready3 <- dat_ready3 %>%
  mutate(
    env = factor(env),
    geno = factor(geno),
    curve_id = factor(curve_id),
    time = as.numeric(time))

set.seed(1)
mpm <- r4pde::compare_curves(
  data          = dat_ready3,
  time          = "time",
  response      = "ypm",
  treatment     = "geno",
  environment   = "env",
  cluster_k     = 4,
  show_progress = TRUE,
  perm_unit     = "geno",
  perm_strata   = "env",
  k_smooth = 10,
  k_env = 4,
  k_trt = 4,
  gamma = 1.6)

D_mildew <- mpm$distance

set.seed(1)
mts <- r4pde::compare_curves(
  data          = dat_ready1,
  time          = "time",
  response      = "yts",
  treatment     = "geno",
  environment   = "env",
  cluster_k     = 3,
  show_progress = TRUE,
  perm_unit     = "geno",
  perm_strata   = "env",
  k_smooth = 10,
  k_env = 4,
  k_trt = 4,
  gamma = 1.6)

D_tan <- mts$distance
summary(mts$gam)

set.seed(1)
mlr <- r4pde::compare_curves(
  data          = dat_ready2,
  time          = "time",
  response      = "yrust",
  treatment     = "geno",
  environment   = "env",
  cluster_k     = 4,
  show_progress = TRUE,
  perm_unit     = "geno",
  perm_strata   = "env",
  k_smooth = 10,
  k_env = 4,
  k_trt = 4,
  gamma = 1.6)

D_rust <- mlr$distance
```

## 1. Helper Function: Check and Align Matrices

This function receives several distance matrices and keeps only the cultivars that are present in all of them.

```{r}
align_distance_matrices <- function(...) {
  mats <- list(...)
  
  # Intersection of cultivars
  common_ids <- Reduce(intersect, lapply(mats, rownames))
  
  mats2 <- lapply(mats, function(M) {
    M[common_ids, common_ids, drop = FALSE]
  })
  
  return(mats2)
}
```

## 2. Function: Distance to Resistant Group

For each cultivar, calculate the average distance to the cultivars in the resistant cluster.

```{r}
distance_to_resistant_group <- function(D, clusters, resistant_cluster) {
  
  # Ensure alignment between vector and matrix names
  ids <- intersect(rownames(D), names(clusters))
  D <- D[ids, ids, drop = FALSE]
  clusters <- clusters[ids]
  
  resistant_ids <- names(clusters)[clusters == resistant_cluster]
  
  if(length(resistant_ids) < 1) {
    stop("No cultivar found in the resistant cluster.")
  }
  
  # Average distance to the resistant group
  dist_mean <- sapply(ids, function(g) {
    mean(D[g, resistant_ids], na.rm = TRUE)
  })
  
  # Optional: exclude self-distance if g is also in the resistant group
  dist_mean2 <- sapply(ids, function(g) {
    ref <- resistant_ids
    if(g %in% ref && length(ref) > 1) ref <- setdiff(ref, g)
    if(length(ref) == 0) return(NA_real_)
    mean(D[g, ref], na.rm = TRUE)
  })
  
  tibble(
    genotype = ids,
    dist_to_resistant = dist_mean2
  )
}

```

## 3. Align Matrices

```{r}
aligned <- align_distance_matrices(D_rust, D_mildew, D_tan)

D_rust2   <- aligned[[1]]
D_mildew2 <- aligned[[2]]
D_tan2    <- aligned[[3]]
# Keep only common cultivars in the cluster vectors as well
common_ids <- Reduce(intersect, list(
  rownames(D_rust2),
  names(cl_rust),
  names(cl_mildew),
  names(cl_tan)
))

D_rust2   <- D_rust2[common_ids, common_ids]
D_mildew2 <- D_mildew2[common_ids, common_ids]
D_tan2    <- D_tan2[common_ids, common_ids]

cl_rust   <- cl_rust[common_ids]
cl_mildew <- cl_mildew[common_ids]
cl_tan    <- cl_tan[common_ids]
```

## 4. Define the Resistant Cluster for Each Disease

```{r}
res_cluster_rust   <- 1
res_cluster_mildew <- 1
res_cluster_tan    <- 2
```

## 5. Distance to the Resistant Ideotype by Disease

```{r}
rust_tbl <- distance_to_resistant_group(
  D = D_rust2,
  clusters = cl_rust,
  resistant_cluster = res_cluster_rust) %>%
  rename(dist_rust = dist_to_resistant)

mildew_tbl <- distance_to_resistant_group(
  D = D_mildew2,
  clusters = cl_mildew,
  resistant_cluster = res_cluster_mildew) %>%
  rename(dist_mildew = dist_to_resistant)

tan_tbl <- distance_to_resistant_group(
  D = D_tan2,
  clusters = cl_tan,
  resistant_cluster = res_cluster_tan) %>%
  rename(dist_tan = dist_to_resistant)
```

## 6. Multi-disease Table

```{r}
multi_tbl <- rust_tbl %>%
  left_join(mildew_tbl, by = "genotype") %>%
  left_join(tan_tbl, by = "genotype")
```

## 7. Standardization and Multi-disease Index

*Lower values indicate better multiobjective resistance.*

```{r}
library(dplyr)
multi_tbl <- multi_tbl %>%
  mutate(
    z_rust   = as.numeric(scale(dist_rust)),
    z_mildew = as.numeric(scale(dist_mildew)),
    z_tan    = as.numeric(scale(dist_tan))) %>%
  mutate(
    mean_z = (z_rust + z_mildew + z_tan) / 3,
    sd_z   = apply(dplyr::select(., z_rust, z_mildew, z_tan), 1, sd, na.rm = TRUE),
    index_multi = mean_z + 0.5 * sd_z) %>%
  arrange(index_multi) %>%
  mutate(rank_multi = row_number())

# lower index_multi = better
multi_tbl
```

## 8. Pareto Front

```{r}
is_dominated <- function(i, M) {
  any(apply(M, 1, function(row) {
    all(row <= M[i, ]) && any(row < M[i, ])
  }))
}

M <- as.matrix(multi_tbl[, c("dist_rust", "dist_mildew", "dist_tan")])

dominated <- sapply(seq_len(nrow(M)), is_dominated, M = M)

multi_tbl <- multi_tbl %>%
  mutate(pareto = !dominated)

multi_tbl %>%
  arrange(index_multi) %>%
  dplyr::select(genotype, dist_rust, dist_mildew, dist_tan, index_multi, rank_multi, pareto)


```

# Plots

```{r}
in1 = ggplot(multi_tbl, aes(x = mean_z, y = sd_z, label = genotype, color = pareto)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(size = 3) +
  theme_bw() +
  labs(
    x = "Mean standardized distance",
    y = "Between-disease variability",
    color = "Pareto front") +
  scale_color_manual(
    values = c(
      "TRUE" = "#C11C84",
      "FALSE" = "#f7c841"))



library(scales)

tri_tbl <- multi_tbl %>%
  mutate(
    rust_res   = 1 - rescale(dist_rust),
    mildew_res = 1 - rescale(dist_mildew),
    tan_res    = 1 - rescale(dist_tan)
  ) %>%
  rowwise() %>%
  mutate(
    total = rust_res + mildew_res + tan_res,
    rust_w   = rust_res / total,
    mildew_w = mildew_res / total,
    tan_w    = tan_res / total
  ) %>%
  ungroup() %>%
  mutate(
    x = 0.5 * (2 * mildew_w + tan_w),
    y = (sqrt(3) / 2) * tan_w
  )

triangle <- tibble(
  x = c(0, 1, 0.5, 0),
  y = c(0, 0, sqrt(3)/2, 0)
)





in2 = ggplot() +
  geom_polygon(data = triangle, aes(x, y), fill = NA, color = "black") +
  geom_point(data = tri_tbl, aes(x, y, color = index_multi, shape = pareto), size = 3) +
  ggrepel::geom_text_repel(data = tri_tbl, aes(x, y, label = genotype), size = 3) +
  annotate("text", x = 0, y = -0.03, label = "Rust", hjust = 0.5) +
  annotate("text", x = 1, y = -0.03, label = "Mildew", hjust = 0.5) +
  annotate("text", x = 0.5, y = sqrt(3)/2 + 0.04, label = "Tan spot", hjust = 0.5) +
  theme_void()

in2 = ggplot() +
  geom_polygon(data = triangle, aes(x, y), fill = NA, color = "black") +
  geom_point(data = tri_tbl, aes(x, y, color = index_multi, shape = pareto), size = 3) +
  ggrepel::geom_text_repel(data = tri_tbl, aes(x, y, label = genotype), size = 3) +
  annotate("text", x = 0, y = -0.03, label = "Leaf Rust", hjust = 0.5) +
  annotate("text", x = 0.95, y = -0.04, label = "Powdery\nMildew", hjust = 0.5) +
  annotate("text", x = 0.5, y = sqrt(3)/2 + 0.04, label = "Tan spot", hjust = 0.5) +
  scale_color_gradientn(colors = c("#C11C84", "#f7c841"),
                        name = "Multi-disease index") +
  coord_cartesian(clip = "off") +
  theme_void() +
  theme(plot.margin = margin(10, 40, 20, 10))+
  labs(shape = "Pareto front")

(in1|in2)+plot_annotation(tag_levels = "A")
# ggsave("multi.png", width = 16, height = 8
#       , bg = "white", dpi = 1000)
```

```{r}
mix_cols <- function(w1, w2, w3,
                     col1 = "#C11C84",   # Leaf Rust
                     col2 = "#f7c841",   # Powdery Mildew
                     col3 = "#3A7CA5") { # Tan spot
  
  rgb1 <- col2rgb(col1)
  rgb2 <- col2rgb(col2)
  rgb3 <- col2rgb(col3)
  
  r <- w1 * rgb1[1, ] + w2 * rgb2[1, ] + w3 * rgb3[1, ]
  g <- w1 * rgb1[2, ] + w2 * rgb2[2, ] + w3 * rgb3[2, ]
  b <- w1 * rgb1[3, ] + w2 * rgb2[3, ] + w3 * rgb3[3, ]
  
  rgb(r, g, b, maxColorValue = 255)}

tri_tbl <- tri_tbl %>%
  mutate(
    point_col = mix_cols(rust_w, mildew_w, tan_w))

edge_gradient <- function(p1, p2, col1, col2, n = 200) {
  t_vals <- seq(0, 1, length.out = n)
  
  x_vals <- (1 - t_vals) * p1[1] + t_vals * p2[1]
  y_vals <- (1 - t_vals) * p1[2] + t_vals * p2[2]
  
  tibble::tibble(
    x = x_vals[-n],
    y = y_vals[-n],
    xend = x_vals[-1],
    yend = y_vals[-1],
    col = grDevices::colorRampPalette(c(col1, col2))(n - 1)
  )
}

v_rust <- c(0, 0)
v_mild <- c(1, 0)
v_tan  <- c(0.5, sqrt(3)/2)

edge_rm <- edge_gradient(v_rust, v_mild, "#C11C84", "#f7c841")
edge_rt <- edge_gradient(v_rust, v_tan,  "#C11C84", "#3A7CA5")
edge_mt <- edge_gradient(v_mild, v_tan,  "#f7c841", "#3A7CA5")

edges <- dplyr::bind_rows(edge_rm, edge_rt, edge_mt)
tri_tbl <- tri_tbl %>%
  arrange(genotype) %>%  # opcional (ordem alfabética)
  mutate(id = as.numeric(factor(genotype)))

in3 = ggplot() +
  geom_segment(
    data = edges,
    aes(x = x, y = y, xend = xend, yend = yend, color = col),
    linewidth = 2,
    lineend = "round",
    show.legend = FALSE) +
  scale_color_identity() +
  geom_point(
    data = tri_tbl,
    aes(x, y, shape = pareto),
    size = 2.5,
    color = tri_tbl$point_col,
    position = position_jitter(width = 0.01, height = 0.01)) +
  ggrepel::geom_text_repel(
    data = tri_tbl,
    aes(x, y, label = genotype),
      size = 3,
  box.padding = 0.3,
  point.padding = 0.1,
  force = 0.5,
  segment.color = "gray40",   # 👈 linha
  segment.size = 0.3,
  segment.alpha = 0.7,
   min.segment.length = 0) +
  annotate("text", x = 0, y = -0.03, label = "Leaf Rust",
           hjust = 0.5, color = "#C11C84") +
  annotate("text", x = 0.95, y = -0.04, label = "Powdery\nMildew",
           hjust = 0.5, color = "#f7c841") +
  annotate("text", x = 0.5, y = sqrt(3)/2 + 0.04, label = "Tan Spot",
           hjust = 0.5, color = "#3A7CA5") +
  coord_cartesian(clip = "off") +
  theme_void() +
  theme(plot.margin = margin(10, 40, 20, 10))+
     labs(shape = "Pareto front")
 in3
 (in1/in3)+plot_annotation(tag_levels = "A")
 
# ggsave("multi2.png", width =9, height = 12
#       , bg = "white", dpi = 1000)

```

```{r}
library(dplyr)
library(ggplot2)
library(ggrepel)
library(patchwork)

## REPLACE with the actual distance column names
## example:
## rust = dist_rust
## mildew = dist_mildew
## tan = dist_tan

plot_pareto_2d <- function(dat, xvar, yvar, xlab, ylab) {
  
  x <- rlang::ensym(xvar)
  y <- rlang::ensym(yvar)
  
  ## keep only Pareto points
  pf <- dat %>%
    filter(pareto == "Pareto" | pareto == 1 | pareto == TRUE) %>%
    arrange(!!x, !!y)
  
  ggplot(dat, aes(x = !!x, y = !!y)) +
    geom_point(aes(shape = pareto), color = dat$point_col, size = 3.5) + 
    labs(x = xlab, y = ylab, shape = "Pareto") +
    theme_classic()
}

p_rm <- plot_pareto_2d(
  tri_tbl,
  dist_rust, dist_mildew,
  "Leaf Rust distance",
  "Powdery Mildew distance"
)

p_rt <- plot_pareto_2d(
  tri_tbl,
  dist_rust, dist_tan,
  "Leaf Rust distance",
  "Tan Spot distance"
)

p_mt <- plot_pareto_2d(
  tri_tbl,
  dist_mildew, dist_tan,
  "Powdery Mildew distance",
  "Tan Spot distance"
)

(p_rm | p_rt | p_mt) + plot_annotation(tag_levels = "A")
```
